\propertitle

### 2019-11-26 maxNt_v6.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-11-25T19:19:26+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal population\\ from subpopulation data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

%\newcommand*{\yXv}{\varSigma}
\newcommand*{\yAv}{A}
%\newcommand*{\yxv}{s}
\newcommand*{\yav}{a}
\newcommand*{\yNv}{N}
\newcommand*{\yNN}{\varNu}
\newcommand*{\yFF}{F}
\newcommand*{\yff}{f}
\newcommand*{\yF}{\bm{\yFF}}
\newcommand*{\yf}{\bm{\yff}}
\newcommand*{\yFFm}{\yFF^{*}}
\newcommand*{\yFm}{\yF^{*}}

@@

What correlations are important for the description of the multi-neuronal
activity in a specific brain area? How does such activity change when
external stimuli or experimental conditions change? Does such activity
range over all its mathematically possible values, or only over a
subset thereof?

Answering this kind of questions always engages an element of uncertainty.
We can't say \enquote{the answer is such and such}, but can at best assign
a degree of reasonable belief -- that is, probability -- to every possible
answer. The assessment of this distribution of belief involves experimental
data, such as recordings of neuronal activity from specific brain areas,
and pre-data knowledge about biological conditions and mechanisms. Our
pre-data \dobs\ are often simplified to be mathematically more tractable,
and are therefore called \enquote{models}.

Despite remarkable advances in recording technologies, the best
experimental measurements of instantaneous neuronal activity can still only
record a very small sample of neurons -- hundreds at most -- compared to
the numbers that constitute a functionally distinguished brain region. Many
probabilistic models focus on such samples only, somehow neglecting, in
their assumptions, that the recorded neurons are a sample from a larger
population. This kind of isolation assumptions sometimes escape attention,
being subtly hidden in the mathematics. Some probabilistic models try to
take also unrecorded neurons into account individually, and therefore
become very complex\mynote{Yasser \amp\ Vahid: refs for here?}. It would be
useful to have models that operate in between: addressing the larger brain
area whence the sample comes, but without adding too much complexity and
detail about it. Such intermediate models would be useful, for example, for
preliminary investigations, to help us to decide which hypotheses to
discard and which to consider for more complex and costly studies.

\mynote{[Yasser:] In the present work we aim to address the issue of building probability distributions over a large population using recording from subpopulations by considering the distribution of total population activity as an example. [}
In the present work we propose such an intermediate model. It answers this
question:\mynote{]} How much can the \emph{total} activity of a large neuronal
population have been, given the observation of the activity of a very small
sample thereof? This model addresses a larger brain area, avoiding the
assumption of isolation of the sample; and by focusing on the total
activity, rather than the activity of individual neurons, it remains simple
and numerically tractable.

The model we propose is based on the straightforward combination of the
maximum-entropy method and basic sampling relations from the probability
calculus, discussed in \sect~\ref{sec:model}. The maximum-entropy or
minimum-relative-entropy method \citep{jaynes1957}[much clearer
in][]{jaynes1963}{sivia1996_r2006,hobsonetal1973,jaynes1985b_r1996,grandy1980}
has been used for different kinds of estimations of the neuronal activity
of various brain areas and about other phenomena of importance to the
neurosciences, for example gene and protein interaction \citep[for
example][]{martignonetal1995,bohteetal2000,shlensetal2006,schneidmanetal2006,tkaciketal2006,mackeetal2009b,tkaciketal2009,roudietal2009c,barreiroetal2010,gerwinnetal2010,mackeetal2011b,ganmoretal2011,cohenetal2011,granotatedgietal2013,mackeetal2013,tkaciketal2014b,shimazakietal2015,moraetal2015,lezonetal2006,weigtetal2009}.

We illustrate possible uses of our proposed model in
\sect~\ref{sec:application}, by applying it to two concrete data sets:
\begin{enumerate*}[label=(\alph*)]
\item the activity of 65 neurons recorded for 20 min from a rat's Medial
  Entorhinal Cortex \citep{stensolaetal2012}, \item the activity of 159
  neurons recorded for 15 min from a macaque's Motor Cortex
  \citep{rostamietal2016_r2017}.
\end{enumerate*}
For each data set the model gives the most plausible frequency distribution
with which all levels of total activity of a larger population of 1\,000--10\,000
neurons appeared during the recording. These guessed frequency
distributions are distinctly different from the recorded ones for the
sampled neurons. For the first data set, for example, the frequency
distribution for the full population has two very distinct modes, both at low
activities, %($0.7\%$ and $2.6\%$ of simultaneously active neurons
% within $3\;\mathrm{ms}$,
see \fig~\ref{fig:plots_5mom}, whereas the frequency distribution for the
sample is monotonically decreasing with its maximum at zero activity. The
frequency distribution for the second data set doesn't have two modes but
still presents one prominent shoulder in its low-activity mode. Note that
these guessed features of the full population cannot be inferred by the
application of maximum-entropy \emph{to the sample alone}. Although we don't
formulate any hypotheses on why the larger population could have two distinct
low-activity regimes -- the purpose of the present work is only
methodological -- these results shows that the proposed method can lead to
the formulation or preliminary assessment of interesting hypotheses.

We want to stress the usefulness of making a quantified guess about the
activity of a larger brain area. Such a guess seems indeed to be the
primary idea behind recording a sample from that area. %  Second, it allows us
% to make comparisons across experimental sessions; such comparisons would me
% difficult or meaningless if made with the recorded samples, which generally
% comprise non-overlapping sets of neurons and differ in size.
There's also an important methodological reason. Maximum-entropy methods
have been used to assess the informational sufficiency of pairwise
correlations and correlations of higher order \citep[see for
example][]{martignonetal1995,bohteetal2000,schneidmanetal2006,shlensetal2006,barreiroetal2010,ganmoretal2011,granotatedgietal2013}.
But when the difference in size between sample and full population is too
large, correlation sufficiency for the sample implies the \emph{lack} of
correlation sufficiency for the larger population, and vice versa. Therefore,
maximum-entropy applications at the sample level can be deceptive for
questions of statistical sufficiency, and a full-population application is
more reliable. We discuss this point in detail in \sect~\ref{sec:marginalization}.

@@

We obviously don't know whether the frequency distribution obtained with
our approach is the \emph{actual} one which the activity levels of the full
population had during the recording; it's only the most plausible. In high
dimensions, however, the features of the most plausible distribution may
\emph{not} be typical of the majority of most plausible distributions; and
the set of all possible frequency distributions, if the full population for
example comprises 1\,000 neurons, is a 1\,000-dimensional space. In
\sect~\ref{sec:assumptions} we therefore try to assess which features of
the frequency distribution delivered by our method may be typical and
therefore expected of the actual one. We find that general features such as
the bimodality of the first data set are indeed typical. Maximum-entropy
models can be considered as approximations of Bayesian models based on
various assumptions of inferential sufficiency
\citep{jaynes1986d_r1996,portamana2017}. How do our guesses change if we
modify our pre-data assumptions? We show, in the same section, that the
typical features indicated by our method are robust against such changes.

@@

\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/newpaper3_gridvsnogrid.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/newpaper3_gridconv.pdf}%
\caption{***}
\label{fig:plots_cellsubsets}
\end{figure}
% maxent_solutions_different_moments_partial.nb
These results suggest that this maximum-entropy approach could be very
useful. Here is a toy example of possible uses. We could be interested in
the hypothesis that two distinct cell types or assemblies be present in the
region where the recording was made. Finding a full-population distribution
with two peaks, as in the upper panel of \fig~\ref{fig:plots_5mom}, would
provide some evidence for this hypothesis. Let's further imagine that we
have reasons for suspecting that a specific set of the sampled neurons are
of one type, and the remaining of the other type. In the case of the first
data set, 27 of the 65 sampled neurons were identified as grid cells
belonging to 3--4 functional modules, see \cite{dunnetal2015}. Could the
two peaks in the distribution of \fig~\ref{fig:plots_5mom} reflect the
activities of grid versus non-grid cells? We apply the method to these two
sets of neurons individually, using $N=4\,150$ for the grid set and
$N=5\,850$ for the non-grid set, to reflect their proportions in the
recorded sample. The results are shown in the upper panel of
\fig~\ref{fig:plots_cellsubsets}. The distribution for the larger
population of grid cells (\textcolor{mygreen}{green triangles}) seems to
have one, broad peak, close to the first peak of the full-population
distribution. The distribution for the larger population of non-grid cells
(\textcolor{myred}{red circles}) has two peaks instead, roughly at the same
positions as the full-population distribution (\textcolor{myblue}{blue
  curve}) but closer in height. So it would seem that the population of
grid cells is contributing to the first peak of the full population, but it
isn't the sole contributor. We can also assess if the distributions for the
two sets are independent. If they were independent, the full-population
distribution would be given by their convolution:
\begin{equation}
  \label{eq:convolution}
  P_{\text{full}}(\yA) =  \smashoperator{\sum_{\yA'=\max(0,\yA-5850)}^{\min(\yA,4150)}} P_{\text{grid}}(\yA')\;
  P_{\text{non-grid}}(\yA-\yA'). %, \qquad \yA \in \set{1,\dotsc,10\,000}
\end{equation}
However, the lower panel of
figure~\ref{fig:plots_cellsubsets} shows that such convolution
(\textcolor{black}{black diamonds}) is different from the full-population
distribution (\textcolor{myblue}{blue curve}): the two peaks of the former
are closer in position and height. This means that the population
distributions of grid and non-grid cells are \emph{not} independent:
knowledge of the activity of either set gives us some information about the
activity of the other.

@@

The toy analysis just given should be taken not literally, but just as an
illustration of the method's possible applications. The important point is
that this method is computationally very cheap but can provide useful
insights, even if just qualitatively.

@@

First of all the sampling relation. We have a set of $N$ units, $\yA$ of
which have some specific property, and we sample in an unknown way $n$ of
the $N$ units. The probability that $\ya$ of the $n$ sampled units have that
property is then given by the hypergeometric distribution
\begin{equation}
  \label{eq:conditional_hypergeometric}
  \yG_{\ya\yA} \defd \pf(\ya \|\yA, n,N)=
  \binom{n}{n\ya}\binom{\yNv-n}{\yNv \yA-n\ya}\binom{\yNv}{\yNv \yA}^{-1}.
\end{equation}
typical of \enquote{drawing without replacement}
\citep[\chap~3]{jaynes1994_r2003}[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968}.
In the following leave $n$, $N$ implicit in the conditional. If we are
uncertain about the number $\yA$, with belief $P(\yA)$, then by the theorem
of total probability our belief about $\ya$ is
\begin{equation}
  \label{eq:belief_sample_unknown_A}
  p(\ya)= \sum_{\yA} \yG_{\ya\yA} \;P(\yA).
\end{equation}
In our case the units are neurons, the set is the full population, and the
property is their being active in a specific time bin.

@@

We have already introduced the notation for the sample neurons. We
introduce an analogous notation for the $\yNv$ neurons constituting the
larger population, but using the corresponding capital letters:
$\yAv_{i}(t)$ is the activity of the $i$th neuron at time bin $t$,
$\yA(t) \defd \sum_{i} \yAv_{i}(t)/\yNv$ is the activity at that
bin averaged over the larger population, and so on.

@@

If $\yHc$ denotes our state of knowledge -- the evidence and assumptions
backing our probability assignments -- our uncertainty about the full
activity of the larger population is expressed by the joint probability
distribution
\begin{equation}
  \label{eq:joint_plaus}
  \pf(\yAv_1, \yAv_2, \dotsc, \yAv_\yNv \| \yHc)
  \quad\text{or}\quad
\pf(\yAA \| \yHc), \quad \yAA \in \set{0,1}^\yNv.
\end{equation}
Our uncertainty about the state of the sample is likewise expressed by
\begin{equation}
  \label{eq:sample_plaus}
  \pf(\yav_1, \yav_2, \dotsc, \yav_n \| \yHc) \quad\text{or}\quad
\pf(\yaa \| \yHc), \quad \yaa \in \set{0,1}^n.
\end{equation}

@@

The \emph{representation theorem for finite exchangeability} states that,
in the state of knowledge $\yH$, the symmetric distribution for the full
activity is completely determined by the distribution for its
population-average:
\begin{equation}
  \label{eq:joint_plaus_N_homog}
  \pf(\yAA \|  \yH) \equiv
  \sum_{\yA}\pf(\yAA \| \yA, \yH)\,
  \pf(\yA \| \yH) =
  \binom{\yNv}{\yNv\yA}^{-1} \pf(\yA \| \yH).
\end{equation}
The equivalence on the left is just an application of the law of total
probability; the equality on the right is the statement of the theorem.
This result is intuitive: owing to symmetry, we must assign equal
probabilities to all $\tbinom{\yNv}{\yNv\yA}$ activity vectors with
$\yNv\yA$ active neurons; the probability of each activity vector is
therefore given by that of the average activity divided by the number of
possible vector values. Proof of this theorem and generalizations to
non-binary and continuum cases are given by de~Finetti
\citey{definetti1959b}, Kendall \citey{kendall1967}, Ericson
\citey{ericson1976}, Diaconis \amp\ Freedman
\citey{diaconis1977,diaconisetal1980}, Heath \amp\ Sudderth
\citey{heathetal1976}.

Our uncertainties about the full population and the sample are connected
via the conditional probability
\begin{equation}
  \label{eq:conditional_hypergeometric}
  \pf(\ya \|\yA, \yH)=
  \binom{n}{n\ya}\binom{\yNv-n}{\yNv \yA-n\ya}\binom{\yNv}{\yNv \yA}^{-1}
  \defs \yG_{\ya\yA},
\end{equation}
which is a hypergeometric distribution, typical of \enquote{drawing without
  replacement} problems. The combinatorial proof of this expression is in
fact the same as for this class of problems
\citep[\chap~3]{jaynes1994_r2003}[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968}.

@@

In particular, when the function $\yg$ is the $m$-activity of the sample,
$\yg_{\ya} = \sav{\yaa\dotso\yaa} \equiv \binom{n\ya}{m}/\binom{n}{m}$, we
find
\begin{multline}
  \label{eq:expe_products}
%  \label{eq:pullback_m_expectations}
%  \yG^*\colon \sav{\underbrace{\yx \dotsm \yx}_{\text{$m$ factors}}}
%  \mapsto
    % \bigl( \sav{\underbrace{\yx \dotsm \yx}_{\text{$m$ factors}}}\bigr)^* =
    % \av{\underbrace{\yX \dotsm \yX}_{\text{$m$ factors}}},
    % \\[2\jot]
    % \expe{\sav{\underbrace{\yx \dotsm \yx}_{\text{$m$ factors}}} \| \yH} =
    % \expe{\av{\underbrace{\yX \dotsm \yX}_{\text{$m$ factors}}} \| \yH}
  \E(\sav{\underbrace{\yaa \dotsm \yaa}_{\text{$m$ factors}}} \| \yH)
\equiv
    % \\[2\jot]
    % {}
    % \frac{(n-m)!}{n!}
    \sum_{\ya} %m!\,
    \binom{n}{m}^{-1}
    \binom{n \ya}{m}\, \pf(\ya \| \yH)
    % \sum_{N \yA=0}^N %m!\,
    % \binom{N-m}{N \yA-m} \binom{N}{N \yA}^{-1} \p(\yXf=\yA \| \yH)
    ={}\\
    \binom{\yNv}{m}^{-1}
    % \frac{(N-m)!}{N!}
    \sum_{\yA} %m!\,
    \binom{\yNv \yA}{m}\, \pf(\yA \| \yH)
\equiv    \E(\av{\underbrace{\yAA \dotsm \yAA}_{\text{$m$ factors}}} \| \yH),
\end{multline}
that is, \emph{the expected values of the $m$-activities of the sample and
  of the full population are equal}. The proof of the middle equality uses
the expression for the $m$th factorial moment of the hypergeometric
distribution and can be found in \textcite{potts1953}. Similar relations
can be found for the raw moments $\E(\ya^m)$ and $\E(\yA^m)$, which
can be written in terms of the product expectations using
\eqn~\eqref{eq:products_intermsof_average}.

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-11-26T10:47:14+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal population\\ from subpopulation data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

%\newcommand*{\yXv}{\varSigma}
\newcommand*{\yAv}{A}
%\newcommand*{\yxv}{s}
\newcommand*{\yav}{a}
\newcommand*{\yNg}{N_{\text{g}}}
\newcommand*{\yNng}{N_{\text{ng}}}
\newcommand*{\yNN}{\varNu}
\newcommand*{\yFF}{F}
\newcommand*{\yff}{f}
\newcommand*{\yF}{\bm{\yFF}}
\newcommand*{\yf}{\bm{\yff}}
\newcommand*{\yFFm}{\yFF^{*}}
\newcommand*{\yFm}{\yF^{*}}

@@

What correlations are dominant in the neuronal activity of a specific brain
area? How does such activity change when external stimuli or the activity
of other areas change? Does such activity range over all its mathematically
possible values, or only within restricted bounds?

Answering this kind of questions always engages an element of uncertainty.
We cannot say \enquote{the answer is such and such}; at best we can assign
degrees of reasonable belief -- that is, probabilities -- to the possible
answers. The assessment of this distribution of belief involves
experimental data, such as recordings of neuronal activity from specific
brain areas, and pre-data knowledge about biological conditions and
mechanisms. Our pre-data \dobs\ are often simplified, to be mathematically
tractable, and are therefore called \enquote{models}.

The best experimental measurements of instantaneous neuronal activity use
remarkable technologies, but can still only record a very small sample of
neurons -- hundreds at most -- compared to the numbers that constitute a
functionally distinguished brain region. Many probabilistic models focus on
such samples only: they somehow neglect, in their mathematical assumptions,
that the recorded neurons are a sample from a larger population. Such
isolating assumptions sometimes escape attention, being subtly hidden in
the mathematics. Some probabilistic models try to take unrecorded neurons
into account -- but by describing each neuron individually, thus becoming
very complex \citep{huang2015,battistinetal2017}. It would be useful to
explore models that operate in between, addressing the larger brain area
whence the sample comes, but collectively, without asking about individual
neuronal details. Such intermediate models would be useful for preliminary
investigations, to help us to decide which hypotheses to discard and which
to consider for more complex and costly studies, or to suggest new
hypotheses.

\mynote{[Yasser:] In the present work we aim to address the issue of
  building probability distributions over a large population using
  recording from subpopulations by considering the distribution of total
  population activity as an example. [} In the present work we propose such
an intermediate probability model. It answers this question: How much was
the \emph{total} activity of a large neuronal population, given the
observation of the activity of a very small sample thereof? This model
addresses a larger brain area, avoiding the assumption of isolation of the
sample; and by focusing on the total activity, rather than the activity of
individual neurons, it remains simple and numerically tractable.

The model we propose is based on the straightforward combination of the
maximum-entropy method and basic sampling relations from the probability
calculus, discussed in \sect~\ref{sec:model}. The maximum-entropy or
minimum-relative-entropy method \citep{jaynes1957}[much clearer
in][]{jaynes1963}{sivia1996_r2006,hobsonetal1973,jaynes1985b_r1996,grandy1980}
has been used for different kinds of investigations about the neuronal
activity of various brain areas and about other phenomena of importance to
the neurosciences, for example gene and protein interaction \citep[for
example][]{martignonetal1995,bohteetal2000,shlensetal2006,schneidmanetal2006,tkaciketal2006,mackeetal2009b,tkaciketal2009,roudietal2009c,barreiroetal2010,gerwinnetal2010,mackeetal2011b,ganmoretal2011,cohenetal2011,granotatedgietal2013,mackeetal2013,tkaciketal2014b,shimazakietal2015,moraetal2015,lezonetal2006,weigtetal2009}.

We illustrate possible uses of our proposed model in
\sect~\ref{sec:application}, by applying it to two concrete data sets:
\begin{enumerate*}[label=(\alph*)]
\item the activity of 65 neurons recorded for 20 min from a rat's Medial
  Entorhinal Cortex \citep{stensolaetal2012}, \item the activity of 159
  neurons recorded for 15 min from a macaque's Motor Cortex
  \citep{rostamietal2016_r2017}.
\end{enumerate*}
For each data set the model gives us the most plausible frequencies with
which all levels of total activity of a larger population of
1\,000--10\,000 neurons appeared during the recording. These guessed
frequency distributions are distinctly different from the recorded ones for
the sampled neurons. For the first data set, for example, the frequency
distribution for the larger population has two very distinct modes, both at
low activities %($0.7\%$ and $2.6\%$ of simultaneously active neurons
% within $3\;\mathrm{ms}$,
(see \fig~\ref{fig:plots_5mom}), whereas the frequency distribution for the
sample is monotonically decreasing with its maximum at zero activity. The
frequency distribution for the second data set presents one prominent
shoulder in its low-activity mode. These results show that the proposed
method can lead to the formulation or preliminary assessment of interesting
hypotheses, as we will illustrate with a toy example. Note that these
guessed features of the full population could not be inferred by the
application of maximum-entropy \emph{to the sample alone}.

% We want to stress the usefulness of making a quantified guess about the
% activity of a larger brain area. Such a guess seems indeed to be the
% primary idea behind recording a sample from that area. %  Second, it allows us
% % to make comparisons across experimental sessions; such comparisons would me
% % difficult or meaningless if made with the recorded samples, which generally
% % comprise non-overlapping sets of neurons and differ in size.
% There's also an important methodological reason.

Our approach also solves a methodological problem in the use of
maximum-entropy methods to assess the \enquote{cooperativity},
\enquote{interaction}, or \enquote{synchrony} in neuronal activity, for
example studying its pairwise correlations and correlations of higher order
\citep[see for
example][]{martignonetal1995,bohteetal2000,schneidmanetal2006,shlensetal2006,barreiroetal2010,ganmoretal2011,granotatedgietal2013}.
When the difference in size between a large population and a sample thereof
is too large, the presence of some inferential properties of correlations
for the sample implies the \emph{lack} of such properties for the larger
population, and vice versa. Maximum-entropy applications at the sample
level can therefore deceive us in questions regarding the cooperativity of
the larger population. The application to the larger population is more
reliable. We discuss this point in detail in
\sect~\ref{sec:marginalization}.

@@

\mynote{review this}We obviously don't know whether the activity levels of
the larger population really had the frequencies given by the model, during
the recording. The model gives our most plausible guess. In high
dimensions, however, the features of the most plausible distribution may
\emph{not} be typical of the majority of most plausible distributions; and
the set of all possible frequency distributions, if the full population for
example comprises 1\,000 neurons, is a 1\,000-dimensional space. In
\sect~\ref{sec:assumptions} we therefore try to assess which features of
the frequency distribution delivered by our method may be typical and
therefore expected of the actual one. We find that general features such as
the bimodality of the first data set are indeed typical. Maximum-entropy
models can be considered as approximations of Bayesian models based on
various assumptions of inferential sufficiency
\citep{jaynes1986d_r1996,portamana2017}. How do our guesses change if we
modify our pre-data assumptions? We show, in the same section, that the
typical features indicated by our method are robust against such changes.

@@

\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/newpaper3_gridvsnogrid.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/newpaper3_gridconv.pdf}%
\caption{***}
\label{fig:plots_cellsubsets}
\end{figure}
% maxent_solutions_different_moments_partial.nb
These results suggest that this maximum-entropy approach could be very
useful. Here is a toy example of possible uses. We could be interested in
the hypothesis that two distinct cell types or assemblies be present in the
region where the recording was made. Finding a full-population distribution
with two peaks, as in the upper panel of \fig~\ref{fig:plots_5mom}, would
provide some evidence for this hypothesis. Let's further imagine that we
have reasons for suspecting that a specific set of the sampled neurons are
of one type, and the remaining of the other type. In the case of the first
data set, 27 of the 65 sampled neurons were identified as grid cells
belonging to 3--4 functional modules, see \cite{dunnetal2015}. Could the
two peaks in the distribution of \fig~\ref{fig:plots_5mom} reflect the
activities of grid versus non-grid cells? We apply the method to these two
sets of neurons individually, using $\yNg=4\,150$ for the grid set and
$\yNng=5\,850$ for the non-grid set, to reflect their proportions (27/65
and 38/65) in the recorded sample. The results are shown in the upper panel
of \fig~\ref{fig:plots_cellsubsets}. The distribution for the larger
population of grid cells (\textcolor{mygreen}{green triangles}) seems to
have one, broad peak, close to the first peak of the full-population
distribution. The distribution for the larger population of non-grid cells
(\textcolor{myred}{red circles}) has two peaks instead, roughly at the same
normalized activities as the peaks of the full-population distribution
(\textcolor{myblue}{blue curve}) but closer in height. It would thus seem
that the population of grid cells is contributing to the first peak of the
full population, but it is not the sole contributor. We can also assess if
the distributions for the two sets of neurons are independent. If they were
independent, the full-population distribution would be given by their
convolution:
\begin{equation}
  \label{eq:convolution}
  P_{\text{full}}(\yA) = \sum_{\yA'} P_{\text{grid}}(\yA')\;
  P_{\text{non-grid}}(\yA-\yA') %, \qquad \yA \in \set{1,\dotsc,10\,000}
\end{equation}
where the index $\yA'$ runs from $\max(0,\yA-\yNng)$ to $\min(\yA,\yNg)$.
But this is not the case: the lower panel of
figure~\ref{fig:plots_cellsubsets} shows that such convolution
(\textcolor{black}{black diamonds}) is quite different from the
full-population distribution (\textcolor{myblue}{blue curve}): the two
peaks of the former are closer in position and height than those of the
latter. The population distributions of grid and non-grid cells are
therefore \emph{not} independent: knowledge of the activity of either set
gives us some information about the activity of the other.

@@

The toy analysis just given should be taken not literally, but just as an
illustration of the method's possible applications. The important point is
that this method is computationally very cheap and yet can provide useful
insights, even if just qualitative ones.

@@

First of all the sampling relation. We have a set of $N$ units, $\yA$ of
which have some specific property, and we sample in an unknown way $n$ of
the $N$ units. The probability that $\ya$ of the $n$ sampled units have that
property is then given by the hypergeometric distribution
\begin{equation}
  \label{eq:conditional_hypergeometric}
  \yG_{\ya\yA} \defd \pf(\ya \|\yA, n,N)=
  \binom{n}{n\ya}\binom{N-n}{N \yA-n\ya}\binom{N}{N \yA}^{-1}.
\end{equation}
typical of \enquote{drawing without replacement}
\citep[\chap~3]{jaynes1994_r2003}[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968}.
In the following leave $n$, $N$ implicit in the conditional. If we are
uncertain about the number $\yA$, with belief $P(\yA)$, then by the theorem
of total probability our belief about $\ya$ is
\begin{equation}
  \label{eq:belief_sample_unknown_A}
  p(\ya)= \sum_{\yA} \yG_{\ya\yA} \;P(\yA).
\end{equation}
In our case the units are neurons, the set is the full population, and the
property is their being active in a specific time bin.

@@

We have already introduced the notation for the sample neurons. We
introduce an analogous notation for the $N$ neurons constituting the
larger population, but using the corresponding capital letters:
$\yAv_{i}(t)$ is the activity of the $i$th neuron at time bin $t$,
$\yA(t) \defd \sum_{i} \yAv_{i}(t)/N$ is the activity at that
bin averaged over the larger population, and so on.

@@

If $\yHc$ denotes our state of knowledge -- the evidence and assumptions
backing our probability assignments -- our uncertainty about the full
activity of the larger population is expressed by the joint probability
distribution
\begin{equation}
  \label{eq:joint_plaus}
  \pf(\yAv_1, \yAv_2, \dotsc, \yAv_N \| \yHc)
  \quad\text{or}\quad
\pf(\yAA \| \yHc), \quad \yAA \in \set{0,1}^N.
\end{equation}
Our uncertainty about the state of the sample is likewise expressed by
\begin{equation}
  \label{eq:sample_plaus}
  \pf(\yav_1, \yav_2, \dotsc, \yav_n \| \yHc) \quad\text{or}\quad
\pf(\yaa \| \yHc), \quad \yaa \in \set{0,1}^n.
\end{equation}

@@

The \emph{representation theorem for finite exchangeability} states that,
in the state of knowledge $\yH$, the symmetric distribution for the full
activity is completely determined by the distribution for its
population-average:
\begin{equation}
  \label{eq:joint_plaus_N_homog}
  \pf(\yAA \|  \yH) \equiv
  \sum_{\yA}\pf(\yAA \| \yA, \yH)\,
  \pf(\yA \| \yH) =
  \binom{N}{N\yA}^{-1} \pf(\yA \| \yH).
\end{equation}
The equivalence on the left is just an application of the law of total
probability; the equality on the right is the statement of the theorem.
This result is intuitive: owing to symmetry, we must assign equal
probabilities to all $\tbinom{N}{N\yA}$ activity vectors with
$N\yA$ active neurons; the probability of each activity vector is
therefore given by that of the average activity divided by the number of
possible vector values. Proof of this theorem and generalizations to
non-binary and continuum cases are given by de~Finetti
\citey{definetti1959b}, Kendall \citey{kendall1967}, Ericson
\citey{ericson1976}, Diaconis \amp\ Freedman
\citey{diaconis1977,diaconisetal1980}, Heath \amp\ Sudderth
\citey{heathetal1976}.

Our uncertainties about the full population and the sample are connected
via the conditional probability
\begin{equation}
  \label{eq:conditional_hypergeometric}
  \pf(\ya \|\yA, \yH)=
  \binom{n}{n\ya}\binom{N-n}{N \yA-n\ya}\binom{N}{N \yA}^{-1}
  \defs \yG_{\ya\yA},
\end{equation}
which is a hypergeometric distribution, typical of \enquote{drawing without
  replacement} problems. The combinatorial proof of this expression is in
fact the same as for this class of problems
\citep[\chap~3]{jaynes1994_r2003}[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968}.

@@

In particular, when the function $\yg$ is the $m$-activity of the sample,
$\yg_{\ya} = \sav{\yaa\dotso\yaa} \equiv \binom{n\ya}{m}/\binom{n}{m}$, we
find
\begin{multline}
  \label{eq:expe_products}
%  \label{eq:pullback_m_expectations}
%  \yG^*\colon \sav{\underbrace{\yx \dotsm \yx}_{\text{$m$ factors}}}
%  \mapsto
    % \bigl( \sav{\underbrace{\yx \dotsm \yx}_{\text{$m$ factors}}}\bigr)^* =
    % \av{\underbrace{\yX \dotsm \yX}_{\text{$m$ factors}}},
    % \\[2\jot]
    % \expe{\sav{\underbrace{\yx \dotsm \yx}_{\text{$m$ factors}}} \| \yH} =
    % \expe{\av{\underbrace{\yX \dotsm \yX}_{\text{$m$ factors}}} \| \yH}
  \E(\sav{\underbrace{\yaa \dotsm \yaa}_{\text{$m$ factors}}} \| \yH)
\equiv
    % \\[2\jot]
    % {}
    % \frac{(n-m)!}{n!}
    \sum_{\ya} %m!\,
    \binom{n}{m}^{-1}
    \binom{n \ya}{m}\, \pf(\ya \| \yH)
    % \sum_{N \yA=0}^N %m!\,
    % \binom{N-m}{N \yA-m} \binom{N}{N \yA}^{-1} \p(\yXf=\yA \| \yH)
    ={}\\
    \binom{N}{m}^{-1}
    % \frac{(N-m)!}{N!}
    \sum_{\yA} %m!\,
    \binom{N \yA}{m}\, \pf(\yA \| \yH)
\equiv    \E(\av{\underbrace{\yAA \dotsm \yAA}_{\text{$m$ factors}}} \| \yH),
\end{multline}
that is, \emph{the expected values of the $m$-activities of the sample and
  of the full population are equal}. The proof of the middle equality uses
the expression for the $m$th factorial moment of the hypergeometric
distribution and can be found in \textcite{potts1953}. Similar relations
can be found for the raw moments $\E(\ya^m)$ and $\E(\yA^m)$, which
can be written in terms of the product expectations using
\eqn~\eqref{eq:products_intermsof_average}.
